DRAFT \\ From physical principles to classical and quantum \\ Hamiltonian and Lagrangian particle mechanics

### 2016-07-24 2015CompleteDerivation/completeDerivation.tex
More work on time dependent evolution

We now generalize our discussion to include time dependent dynamics. This is needed when the evolution map itself changes in time or when a change of state variable is time dependent (e.g. changing to a moving frame).

To do this we will extend phase space to include the temporal degree of freedom. We'll find that the dynamics is described by relativistic Hamiltonian mechanics in the extended phase space. We'll also find that particle states divide into standard and anti-states depending whether time is aligned or anti-aligned with the affine parameter. Note that no extra assumption is needed, which makes relativistic mechanics not some additional characterization but simply the correct way to describe time dependent motion of invariant density distributions.

@@

The second case is when we perform a transformation for the time parameter $t'=t'(t)$. The equations of motion transform and we have $d_{t}t' d_{t'}q^i = \partial_{p_i} H$. As both $q^i$ and $H$ are independent of time, they cannot be redefined to include $d_{t}t'$. Physically the transformation is introducing fictitious forces that are not conservative, so they cannot be expressed by a Hamiltonian. But this means that we have an ill defined mathematical framework as the notion of determinism and reversibility is not defined in a way that is independent of time transformations. We need a framework that is capable of handling the fictitious forces as well.

The third case is when we perform a state variable transformation $q'=q'(q,t)$ that is time dependent (e.g. $q'=q+vt$). Our composite state distribution $\rho(q)$ becomes $\rho(q',t)$: it is no longer defined at equal time. This means that both our measure $\mu$ and our metric $d\mu$ need to be modified.

@@

\subsection{Extended phase space}

Our goal is to describe maps that take a state at a particular time and transport it to another state at another time. Therefore we want to extend the domain and co-domain of the evolution map to be the space of all particle states at all times.

The first thing to do is to extend our $n$-dimensional $\mathcal{Q}$, the manifold that defines our units, to include the time variable $t$. We call $\mathcal{M}$ such extended space which is a manifold of dimension $n+1$: state variables and the time variable are independent (i.e. any combination of state and time is valid). A time variable is not simply another state variable: it doesn't identify extra states and it's the variable on which we have defined deterministic and reversible evolution. When changing variables, we still need to make sure that deterministic and reversible motion can be defined on the new time variable. Transformations like $t'=q, \; q'=t$ (exchanging time with another state variable) or $t'=\sqrt{t^2 + q^2}, \; q'=\tan^{-1}(t/q)$ (polar coordinates between time and state variable) clearly do not guarantee deterministic and reversible motion over the new time variable. The change of time variable must preserve the time ordering therefore the transformation has to be strictly monotonic between the two time variables. In this sense, the time variable is special.

@@

\begin{prop}\label{classical_extended_phase_space}
	The extended state space, the set of states at all possible times, for the particles of a homogeneous classical material is a hyper-surface of the symplectic manifold formed by a cotangent bundle $T^*\mathcal{M}$ equipped with the canonical two-form $\omega = \sum dq^i \wedge dp_i - dt \wedge dE$.
\end{prop}

>>>>

We now generalize our discussion to include time dependent dynamics. This is needed when the evolution map is not the same at all instants or when state variables depend on time (e.g. changing to a moving frame).

To do this we will redefine the particle state space to include the temporal degree of freedom. We'll find that the dynamics is described by relativistic Hamiltonian mechanics in the extended phase space. We'll also find that particle states divide into standard and anti-states depending whether time is aligned or anti-aligned with the evolution parameter. Note that no extra assumption is needed, which makes relativistic mechanics simply the correct time dependent description of the evolution of an invariant density distribution, and not the description of a different type of system.

@@

The second case is when we perform a transformation for the time parameter $\hat{t}=\hat{t}(t)$. The equations of motion transform and we have $d_{t}\hat{t} d_{\hat{t}}q^i = \partial_{p_i} H$. As both $q^i$ and $H$ are independent of time, they cannot be redefined to include $d_{t}\hat{t}$. Physically the transformation is introducing fictitious forces that are not conservative, so they cannot be expressed by a Hamiltonian. But this means that we have an ill defined mathematical framework as the notion of determinism and reversibility is not defined in a way that is independent of time transformations. We need a framework that is capable of handling the fictitious forces as well.

The third case is when we perform a state variable transformation $\hat{q}=\hat{q}(q,t)$ that is time dependent (e.g. $\hat{q}=q+vt$). Our composite state distribution $\rho(q)$ becomes $\rho(\hat{q},t)$: it is no longer defined at equal time. This means that our measure $\mu$ needs to be modified to define integrals over the time variable.

@@

\subsection{Complete state space}

The first thing to do is to amend our definition of state space to include all states at all times. It may seem like we are extending our definition of state and state space but, at a closer look, we are not. Since defined state as a physical configuration at a particular time, the set of all states (i.e. the state space) is more accurately the set of all configurations at all times. In the previous sections we tacitly simplified the problem by ignoring time, which made it easier to study the time independent case. This is common in physics and engineering but this is also the source of the above problems.

We'll therefore call \emph{complete state space} the set of all configurations at all times and we'll indicate it by $\bar{\mathcal{S}}$. In that space, a state is an element $\mathcal{s} \in \bar{\mathcal{S}}$ while a time evolution is a line $\bar{\mathcal{s}} \subset \bar{\mathcal{S}}$ that passes once at all times. Specifying an evolution map means giving a set of lines such that each state is traversed once and only once. Note that the transported state variables are perfect to label the evolutions: as they do not change during the evolution we can take them to label not just the values at a particular time, but the evolutions themselves.

With this geometric picture in mind, we can understand better the complete state space $\bar{\mathcal{C}}$ for the composite system. An evolution $\hat{\mathcal{c}}$ will be a distribution of material over all configuration at all times. Along a particle state evolution $\bar{\mathcal{s}}$, though, the distribution is constant as all the material is transported deterministically and reversibly. A state space $\mathcal{c}$, then, is a distribution over an hyper-surface that cuts across all the particle state evolutions. If we think it as a distribution over the transported state variables we are guaranteed to cover all initial conditions no matter the hyper-surface.

The first thing to do is to extend our $n$-dimensional $\mathcal{Q}$, the manifold that defines our units, to include the time variable $t$. We call $\mathcal{M}$ such extended space which is a manifold of dimension $n+1$: state variables and the time variable are independent (i.e. any combination of state and time is valid). A time variable is not simply another state variable: it doesn't identify extra states and it's the variable on which we have defined deterministic and reversible evolution. When changing variables, we still need to make sure that deterministic and reversible motion can be defined on the new time variable. Transformations like $\hat{t}=q, \; \hat{q}=t$ (exchanging time with another state variable) or $\hat{t}=\sqrt{t^2 + q^2}, \; \hat{q}=\tan^{-1}(t/q)$ (polar coordinates between time and state variable) clearly do not guarantee deterministic and reversible motion over the new time variable. To be meaningful, the change of time variable must preserve the time ordering therefore the transformation has to be strictly monotonic between the two time variables. In this sense, the time variable is special.

@@

\begin{prop}\label{classical_extended_phase_space}
	The complete state space $\hat{\mathcal{S}}$ for the particles of a classical material is a hyper-surface of the symplectic manifold formed by a cotangent bundle $T^*\mathcal{M}$ equipped with the canonical two-form $\omega = \sum dq^i \wedge dp_i - dt \wedge dE$.
\end{prop}
